Formalizing the Edmonds-Karp Algorithm

### 2016-03-09 formalization/document/root.tex
Cleanup, building proof document

% Markierungen: TODO
\documentclass{llncs}

\usepackage{etex}
\pagestyle{plain} % turn on page numbers
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
% \usepackage{mathpartir}
%\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{multirow}
\usepackage{pgfplots}

\usepackage{packages/isabelle}
\usepackage{packages/isabelletags}
\usepackage{packages/isabellesym}
\usepackage{packages/comment}

% \isabellestyle{it}

\def\isachardoublequote{}%
\def\isachardoublequoteopen{}%
\def\isachardoublequoteclose{}%

\newcommand{\isainnerkeyword}[1]{{\tt #1}}
\newcommand{\isasymexistsA}{\isamath{\exists_{\textsc A}\,}}

\def\isadelimproof{}
\def\endisadelimproof{}
\def\isatagproof{}
\def\endisatagproof{}
\def\isafoldproof{}
\def\isadelimproof{}
\def\endisadelimproof{}

\input{lstisabelle}
\lstset{basicstyle=\footnotesize\ttfamily\slshape}
\lstset{captionpos=b}
\lstset{numberbychapter=false}

\newcommand{\isai}{\lstinline[language=isabelle,basicstyle=\normalsize\ttfamily\slshape]}

\input{macros}

% Include snippets
\newcommand{\DefineSnippet}[2]{%
   \expandafter\newcommand\csname snippet--#1\endcsname{%
     \begin{quote}
     \begin{isabelle}\footnotesize
     #2
     \end{isabelle}
     \end{quote}}}
\newcommand{\Snippet}[1]{\ifcsname snippet--#1\endcsname\csname snippet--#1\endcsname\else\PackageError{}{No snippet '#1' defined.}{}\fi}
\input{snippets/snippets.out}

% \overfullrule=8pt

@@

\title{Formalizing the Edmonds-Karp Algorithm}
% \subtitle{}

\author{Peter Lammich, S.~Reza Sefidgar}

\institute{Technische Universit\"at M\"unchen, \email{\{lammich,sefidgar\}@in.tum.de}}

\maketitle
\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is accessible even without being
an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.

%We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network. 
%The formal proofs are done in Isabelle/HOL. They closely follow a standard textbook proof and are relatively easy 
%to read, even without being an Isabelle expert.

%We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
%Further refinement yields a verified implementation, which is almost as fast as an unverified  
%reference implementation in Java.

% 
% and a verified imperative implementation of it. Our implementation is roughly 2.5 times slower than an unverified  
% reference implementation in Java. 
% 
% Finally, we prove the upper complexity bound of $O(|V||E|^2)$ for the Edmonds-Karp algorithm.
\end{abstract}

@@

While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation. Moreover, this paper is a case study on elegantly formalizing algorithms.

The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL. In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work. The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}.

% Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants. The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}. This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm. Providing such an implementation is specially important, as it could provide the basis for many verfied programs with practical importance.

\section{The Ford-Fulkerson Method}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation by Cormen et al.~\cite{CLRS09}.

% TODO: Possible alternative presentation
%   First define a relaxed network, without the constraints. Define the flow there. Then add network constraints.
%

A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
Moreover, there are two distinct vertices $s,t\in V$, which are called \emph{source} and \emph{sink}.

A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges. 
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the vertices into two sets, such that one set contains 
the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut. The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
It then pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}%
Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is usually stated as the following three statements being equivalent:
\begin{enumerate}
\item $f$ is a maximum flow in a network $G$.
\item there is no augmenting path in the residual graph $G_f$.
\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
\end{enumerate}

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph. There are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

\section{Formalizing the Ford-Fulkerson Method}\label{sec:abs-formalization}
%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
%    Reader should think: Yeah, they did cool stuff!
%
%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
%    
%  Proof of MinCut-MaxFlow: Follows the textbook proof.
%    Uses Isar to even look like a textbook proof, being comprehensible even without running
%      Isabelle/HOL.
%      
%    For example, \isai{augment_flow_presv_cap}.
%      Present formal proof text. Perhaps oppose it to textbook proof?.
%    
%  Abstract Algo looks like pseudocode presented in texbooks.
%    Oppose our algo to textbook pseudocode.
%

%A formal proof contains many technical details that are not presented in an informal proof. While dealing with the technical details in a proof causes distraction, leaving out such details makes it hard to follow the proofs.  . So, readers may use this informal proof as a guide to better understand the relevant thoughts in our formal proof. Moreover using the proof language Isar, we made the formal proof readable even for readers that are not familiar with Isabelle.

In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
us to define named contexts that fix some parameters and assumptions.
For example, the graph theory is developed in the locale \isai{Graph}, which 
fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
\begin{lstlisting}
  locale Graph = fixes c :: "edge => capacity" begin
    definition "E == {(u, v). c (u, v) ~= 0}"
    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
    [...]
\end{lstlisting}
Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.

Networks are based on graphs, and add the source and sink nodes, as well as the network assumptions:
\begin{lstlisting}
  locale Network = Graph + fixes s t :: node
    assumes no_incoming_s: "\<forall>u. (u, s) \<notin> E"
    [...]
\end{lstlisting}
Most theorems presented in this paper are in the context of the \isai{Network} locale.

\subsection{Presentation of Proofs}
% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof. They tend to be inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.

For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL. 
We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.

As an example consider the proof that for a flow $f$ and a residual flow $f'$, the augmented flow $f\augment f'$ is again a valid flow. In particular, one has to show 
that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
``Equation~26.4'' by ``definition of $\augment$'':

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}
% 
In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

\subsection{Presentation of Algorithms}
In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details. As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}. For example, we define the Ford-Fulkerson method as follows:
\Snippet{ford_fulkerson_algo}

% DONE: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked. We leave it to future work to define a set of combinators and appropriate syntax that allows for more concise presentation of pseudocode.

% The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Isabelle Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_correct}

\section{Refinement in Isabelle/HOL}\label{sec:refinement}
After having stated and proved correct an algorithm on the abstract level, the next step is to provide an (efficient) implementation. 
In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.

A natural way to achieve this formally is \emph{stepwise refinement}~\cite{Wirth71}, and in particular \emph{refinement calculus}~\cite{Back78,BaWr98}, which allows us to systematically transform an abstract algorithm into a more concrete one, preserving its correctness.

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
The bind operation fails iff either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m},

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, \isai{m <= m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
also possible results of \isai{m'}. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
means that, if \isai{P} holds, \isai{m} does not fail and all possible results of \isai{m} satisfy \isai{Q}. Note that we provide different recursion constructs
for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
and disappears when joined with other results.

The Isabelle Refinement Framework also supports data refinement. The representation of results can be changed according to a \emph{refinement relation}, 
which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the set of concrete results that are related to an 
abstract result in \isai{m} by \isai{R}. If \isai{m = fail}, then also \isai{\<Down>R m = fail}.

In a typical program development, one first comes up with an initial version \isai{m_0} of
the algorithm and its specification \isai{P,Q}, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions \isai{m$_i$} of the algorithm,
proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of data refinement, one 
gets \isai{P ==> m$_i$ <= \<Down>R$_i$...R$_1$ spec Q}, showing the correctness of the refined algorithm.
If no data refinement is performed, \isai{R$_i$} is set to the identity relation, in which case \isai{\<Down>R$_i$} becomes the identity function.

Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
them down to statements that do not contain monad constructs any more. In many cases, these verification conditions reflect the core idea of the proof precisely.

Monotonicity of the standard combinators also allows for modular refinement: Replacing a part of a program by a refined version
results in a program that refines the original program. This gives us a natural formal model for statements like ``we implement shortest path finding by BFS'', or ``we use arrays to represent the edge labeling''.

% 
% 
% Purpose: Separation of concerns.
% 
% Short intro on Refinement Framework:
% 
% Programs phrased as error-set monad. With order structure. 
% Refinement: Less possible results (refinement ordering). Error is greatest element (refined by everything).
% 
% Building chains of refinement, at the end there should be a deterministic algorithm (yielding exactly one result), which is
% contained (by transitivity) in the specification.
% 
% Standard combinators (bind, recursion, if, etc) are monotonic wrt. refinement ordering, which allows for modularity:
% Replacing part of a program by a refined version, yields a refined program. 
% 
% Data refinement: Changing the representation of data. Relation between concrete and abstract representation. 
% $\Uparrow m$ is biggest concrete program that refines $m$ wrt.\ relation $R$. 
% 
% The Isabelle Refinement Framework provides the nres-monad, and tools to reason about program refinement, 
% the most important one being a VCG.

\section{The Edmonds-Karp Algorithm}\label{sec:edka}
  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.

  Considerably more effort is required to show that the resulting algorithm terminates within $O(VE)$ iterations. 
  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
  on a shortest path itself.
  On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
  or the number of edges that lie on some shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.
  % TODO: Add more explanation here, like: DIstinguish two cases: 1) ... 2) The distance s-t in cf is still the same, but then ...

  Note that Cormen et al.~present the same idea a bit differently: They define an edge of the residual graph being \emph{critical} if it lies on a shortest path such that it will be flipped by augmentation. Then, they establish an upper bound of how often an edge can get critical during the algorithm. Our presentation is more suited for a formal proof, as we can directly construct a measure function from it, \ie a function from flows to natural numbers, which decreases on every iteration and is bounded by $O(VE)$.

  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add multiple reverse edges and removes at least one original edge requires some generalization of the statement.
  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
  edge, the path does no longer exist.

%   We decided to generalize the statement as follows. Assume we have an original graph $cf$ with a shortest augmenting path $p$, and a transformed graph $cf'$ that has been created from
%   $cf$ by adding some flipped edges from $p$ and removing some edges from $p$.
%   Then, we consider an edge $(u,v) \in p$, a path $p_1$ from $s$ to $v$ in the original graph, and a path $p_2$ from $u$ to $t$ in the transformed graph.
%   
%   By induction on the number of reversed edges in $p_2$, we 
%   show that $|p_1|+|p_2| > |p|$. In the induction step, we use the proof idea for the single flipped edge $(v,u)$, and,
%   if $p_2$ contains more flipped edges, we split $p_2$ at the first flipped edge. Then, the initial segment of $p_2$ contains no flipped edges,
%   and thus also is a path in the original graph, and we can apply the induction hypothesis.
%   
%   Proving that augmentation with a shortest augmenting path actually only adds flipped edges, and removes at least one original edge, required some more work,
%   but was straightforward and yielded no surprises.
%   
%   Finally, we phrase the complexity argument as a measure function $m = l*2|E| + n$, where $l$ is the length of the shortest augmenting path,
%   and $n$ is the number of edges that lie on any shortest path. We show that $m$ is decreased by the loop body.
%   Adding a special case for loop termination (there are no augmenting paths), and observing that $m$ is bounded by $2|V||E|$, yields 
%   the desired upper bound on loop iterations. Refinement of the algorithm to add an explicit loop counter, and asserting the upper bound, is then straightforward.

  Having defined the measure function and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one. Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop with a counter, and assert the upper bound on the counter after the loop.

\section{Refinement to Executable Code}\label{sec:executable}
  In the previous section, we have presented our abstract formalization of the Edmonds-Karp algorithm,
  leaving open how to obtain a shortest augmenting path and how to implement the algorithm. 
  In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.

  \subsection{Using Breadth First Search}
  A standard way to find a shortest path in a graph is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the 
  Isabelle Refinement Framework. Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path. For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm. Note that the resulting BFS algorithm is independent, and can be reused for finding shortest paths in other applications.

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage{isabelle,isabellesym}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage[only,bigsqcap]{stmaryrd}
\usepackage{wasysym}

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

% Tweaks
\newcounter{TTStweak_tag}
\setcounter{TTStweak_tag}{0}
\newcommand{\setTTS}{\setcounter{TTStweak_tag}{1}}
\newcommand{\resetTTS}{\setcounter{TTStweak_tag}{0}}
\newcommand{\insertTTS}{\ifnum\value{TTStweak_tag}=1 \ \ \ \fi}

\renewcommand{\isakeyword}[1]{\resetTTS\emph{\bf\def\isachardot{.}\def\isacharunderscore{\isacharunderscorekeyword}\def\isacharbraceleft{\{}\def\isacharbraceright{\}}#1}}
\renewcommand{\isachardoublequoteopen}{\insertTTS}
\renewcommand{\isachardoublequoteclose}{\setTTS}
\renewcommand{\isanewline}{\mbox{}\par\mbox{}\resetTTS}

\renewcommand{\isamarkupcmt}[1]{\hangindent5ex{\isastylecmt --- #1}}

%\newcommand{\isaheader}[1]{\section{#1}}
%\renewcommand{\isamarkupheader}[1]{#1}

\newcommand{\DefineSnippet}[2]{#2}

@@

\title{Formalizing the Edmonds-Karp Algorithm}
\author{Peter Lammich and S.~Reza Sefidgar}
\maketitle

\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is accessible even without being
an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.
\end{abstract}

\clearpage
\tableofcontents

\clearpage

% sane default for proof documents
\parindent 0pt\parskip 0.5ex

@@

While there exists another formalization of the Ford-Fulkerson method in
Mizar~\cite{Lee05}, we are, to the best of our knowledge, the first that verify a
polynomial maximum flow algorithm, prove the polynomial complexity bound, or
provide a verified executable implementation. Moreover, this paper is a case
study on elegantly formalizing algorithms.
% generated text of all theories
\input{session}

\section{Conclusion}\label{sec:concl}
  We have presented a verification of the Edmonds-Karp algorithm, using a
  stepwise refinement approach. Starting with a proof of the Ford-Fulkerson
  theorem, we have verified the generic Ford-Fulkerson method, specialized it to
  the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number
  of outer loop iterations. We then conducted several refinement steps to derive
  an efficiently executable implementation of the algorithm, including a
  verified breadth first search algorithm to obtain shortest augmenting paths.
  Finally, we added a verified algorithm to check whether the input is a valid
  network, and generated executable code in SML. The runtime of our verified
  implementation compares well to that of an unverified reference implementation
  in Java. Our formalization has combined several techniques to achieve an
  elegant and accessible formalization: Using the Isar proof
  language~\cite{Wenzel99}, we were able to provide a completely rigorous but
  still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement
  Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16} allowed us
  to present the Ford-Fulkerson method on a level of abstraction that closely
  resembles pseudocode presentations found in textbooks, and then formally link
  this presentation to an efficient implementation. Moreover, modularity of
  refinement allowed us to develop the breadth first search algorithm
  independently, and later link it to the main algorithm. The BFS algorithm can
  be reused as building block for other algorithms. The data structures are
  re-usable, too: although we had to implement the array representation of
  (capacity) matrices for this project, it will be added to the growing library
  of verified imperative data structures supported by the Sepref tool, such that
  it can be re-used for future formalizations. During this project, we have
  learned some lessons on verified algorithm development: 
  \begin{itemize} 
  \item
  It is important to keep the levels of abstraction strictly separated. For
  example, when implementing the capacity function with arrays, one needs to
  show that it is only applied to valid nodes. However, proving that, e.g.,
  augmenting paths only contain valid nodes is hard at this low level. Instead,
  one can protect the application of the capacity function by an assertion---
  already on a high abstraction level where it can be easily discharged. On
  refinement, this assertion is passed down, and ultimately available for the
  implementation. Optimally, one wraps the function together with an assertion
  of its precondition into a new constant, which is then refined independently.
  \item Profiling has helped a lot in identifying candidates for optimization.
  For example, based on profiling data, we decided to delay a possible
  deforestation optimization on augmenting paths, and to first refine the
  algorithm to operate on residual graphs directly. 
  \item ``Efficiency bugs''
  are as easy to introduce as for unverified software. For example, out of
  convenience, we implemented the successor list computation by \emph{filter}.
  Profiling then indicated a hot-spot on this function. As the order of
  successors does not matter, we invested a bit more work to make the
  computation tail recursive and gained a significant speed-up. Moreover, we
  realized only lately that we had accidentally implemented and verified
  matrices with column major ordering, which have a poor cache locality for our
  algorithm. Changing the order resulted in another significant speed-up.
  \end{itemize} 
  We conclude with some statistics: The formalization consists of
  roughly 8000 lines of proof text, where the graph theory up to the
  Ford-Fulkerson algorithm requires 3000 lines. The abstract Edmonds-Karp
  algorithm and its complexity analysis contribute 800 lines, and its
  implementation (including BFS) another 1700 lines. The remaining lines are
  contributed by the network checker and some auxiliary theories. The
  development of the theories required roughly 3 man month, a significant amount
  of this time going into a first, purely functional version of the
  implementation, which was later dropped in favor of the faster imperative
  version.

  \subsection{Related Work}\label{sec:related_work} 
  We are only aware
  of one other formalization of the Ford-Fulkerson method conducted in
  Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication on
  this formalization except~\cite{Lee05}, which provides a Mizar proof script
  without any additional comments except that it ``defines and proves
  correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of
  graph manipulations''. Moreover, in Lee et al.~\cite{LeRu07}, which is about
  graph representation in Mizar, the formalization is shortly mentioned, and it
  is clarified that it does not provide any implementation or data structure
  formalization. As far as we understood the Mizar proof script, it formalizes
  an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson
  method. Termination is only proved for integer valued capacities. Apart from
  our own work~\cite{La14,NoLa12}, there are several other verifications of
  graph algorithms and their implementations, using different techniques and
  proof assistants. Noschinski~\cite{Nosch15} verifies a checker for
  (non-)planarity certificates using a bottom-up approach. Starting at a C
  implementation, the AutoCorres tool~\cite{Greenaway15,GAK12} generates a
  monadic representation of the program in Isabelle. Further abstractions are
  applied to hide low-level details like pointer manipulations and fixed size
  integers. Finally, a verification condition generator is used to prove the
  abstracted program correct. Note that their approach takes the opposite
  direction than ours: While they start at a concrete version of the algorithm
  and use abstraction steps to eliminate implementation details, we start at an
  abstract version, and use concretization steps to introduce implementation
  details.
